---
title: "VTPEH 6207 - CheckPoint 2"
author: "Ke Wu"
date: "`r Sys.Date()`"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
  toc: true
toc_depth: 1
number_sections: true
urlcolor: blue
---
*3.1*
Background
  Chronic diseases such as obesity, diabetes, and cardiovascular disease act as 
  major public health concerns in New York State. Behavioral risk factors like 
  physical inactivity, poor diet, and smoking significantly contribute to these
  concerns. Analyzing BRFSS data helps explore how health behaviors relate to 
  chronic disease risk and identify vulnerable populations. The motivation of 
  this project is driven by a better understanding on behavioral factors to 
  support more efficient interventions and decisions that improve community 
  health outcomes.
  
Project Question
  What is the relationship between selected behavioral risk factors and chronic 
  disease indicators in adults, in New York State based on 2023 BRFSS survey 
  data?
  
Data Overview
  This dataset comes from the New York State Department of Health (NYSDOH) 
  Behavioral Risk Factor Surveillance System (BRFSS) 2023 survey. It is a 
  telephone survey coordinated nationally by the CDC and administered at the 
  state level. The study population includes non-institutionalized adults aged 
  18 years and older living in New York State. The data represent self-reported 
  health behaviors, chronic disease status, and preventive health practices. 
  The data collection period is 2023.
#loading data
```{r, results='hide'}

# Packages
library("janitor")

# Load file
data_BRFSS=read.csv("/Users/kewu/Desktop/NYSDOH_BRFSS_SurveyData_2023.csv")

# Preview the data
head(data_BRFSS)
names(data_BRFSS)
str(data_BRFSS)
summary(data_BRFSS)

```


  *3.2.1*
  
```{r}
library(dplyr)

var_table <- data.frame(
  Variable = names(data_BRFSS),
  Class = sapply(data_BRFSS, class),
  Type = sapply(data_BRFSS, function(x){
    if(is.numeric(x) | is.integer(x)) "Continuous"
    else if(is.factor(x) | is.character(x)) "Categorical"
    else "Other"
  })
)

# <=10 variable
vars_keep <- c("SEXVAR", "AGE", "_BMI5", "_SMOKER3", "DIABETE4",
               "GENHLTH", "EXERANY2", "INCOME3")

var_table_10 <- var_table %>% filter(Variable %in% vars_keep)

knitr::kable(var_table_10,
             caption = "Table 1. Overview of selected variables from NYSDOH BRFSS 2023.")
```



*3.2.2.1(IMONTH)*
```{r}
library(ggplot2)

x <- data_BRFSS$IMONTH

#mean and median
x_mean <- mean(x, na.rm = TRUE)
x_median <- median(x, na.rm = TRUE)
#mean/median lines and legend 
ggplot(data_BRFSS, aes(x = IMONTH)) +
  geom_histogram(binwidth = 1, boundary = 0.5, color = "black", fill = "blue") +
  geom_vline(aes(xintercept = x_mean, color = "Mean"), linewidth = 1) +
  geom_vline(aes(xintercept = x_median, color = "Median"), linetype = "dashed", linewidth = 1) +
  scale_color_manual(values = c("Mean" = "pink", "Median" = "lightgreen")) +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Histogram of Interview Month (NYSDOH BRFSS 2023)",
    x = "Interview month",
    y = "Number of participants",
    color = "Statistic",
    caption = "Figure 1. Distribution of interview month in the NYSDOH BRFSS 2023 dataset. Solid red line = mean; dashed blue line = median."
  ) +
  theme_minimal()
```

*3.2.2.2*

```{r}
library(ggplot2)

data_BRFSS$month <- as.numeric(data_BRFSS$IMONTH)
data_BRFSS$day <- as.numeric(data_BRFSS$IDAY)


ggplot(data_BRFSS, aes(x = month, y = day)) +
  geom_point(alpha = 0.4, color = "pink") +
  
  scale_x_continuous(breaks = 1:12) +
  scale_y_continuous(breaks = seq(0, 31, 5)) +
  
  labs(
    title = "Scatterplot of Interview Month vs Interview Day (BRFSS 2023)",
    x = "Interview Month (1 = Jan … 12 = Dec)",
    y = "Interview Day of Month",
    caption = "Figure 2. Scatterplot showing distribution of interview timing by month and day in NYSDOH BRFSS 2023 dataset."
  ) +
  
  theme_minimal()
```



*4*
 This document was generated using CHATGTP to debugging the diagram code 
